Population Growth
========================================================

Deterministic Growth
-------------------------
Consider the simple discrete logistic growth model

```{r, echo=1:5}
r = 1
K = 10
n0 = .1
NT = 30
time = 1:NT

#layout(matrix(c(1,2,3,3),2,2,byrow=TRUE))
plot(c(0,K),c(r,0),pch=18,ylim=c(0,r*1.1),
     xlim=c(0,K*1.1),bty='n',xaxt="n", yaxt="n",
     ylab="Per Capita Growth Rate",
     xlab="Population Size",cex.lab=1.5,cex=2,mgp=c(1,1,0)*1.7)
axis(1, pos=0)
axis(2, pos=0)
abline(h=r,lty=3,lwd=2)
abline(a=r,b=-r/K,lwd=3)
text(0.45,0.85,"r",cex=2)
text(10,0.15,"K",cex=2)

plot0 <- function(...){
  plot(...,bty='n',xaxt="n", yaxt="n",cex.lab=1.5,cex=2,mgp=c(1,1,0)*1.8)
  axis(1, pos=0)
  axis(2, pos=0)
}
x = seq(0,K,length=500)
plot0(x,r*x*(1-x/K),lwd=3,type='l',
      xlab="Population Size",
      ylab="Growth Rate"
      )
points(c(0,K),c(0,0),pch=18,cex=2)
abline(a=0,b=r,lty=3,lwd=2)
```

can then simulate as

```{r}
n = rep(n0,NT)
ne = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K)
    ne[t] = ne[t-1] + r*ne[t-1]
}
```


You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot0(time,n,ylim=c(0,12),lwd=3,type='l',
      xlab="Time",ylab="Population Size")
lines(time,ne,lty=3,lwd=3)
text(1,1.3,expression(N[0]),cex=2)
```

Observation Error
-------------------------

One of the realities of science is that we rarely get to observe our system without there being some noise or uncertainty in the observations themselves.
```{r fig.width=7, fig.height=6}
sd.o = 0.25
plot0(time,n,type='l',ylim=c(0,12))
points(time,n+rnorm(NT,0,sd.o))
```

```{r,echo=3:4, fig.width=7, fig.height=6}
ciEnvelope <- function(x,ylo,yhi,...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
  ylo[1])), border = NA,...) 
}
plot0(time,n,type='l',ylim=c(0,12),
     xlab="Time",ylab="Population Size")
ciEnvelope(time,n-1.96*sd.o,n+1.96*sd.o,col="lightGrey")
lines(time,n,lwd=3)
points(time,n+rnorm(NT,0,sd.o))
abline(h=0,v=0)
#lines(time,n+1.96*sd.o,lty=2)
#lines(time,n-1.96*sd.o,lty=2)


```



Observation error does not affect the underlying process itself. The process here is still deterministic, we're just observing it imperfectly


Parameter Uncertainty
-------------------------

We next want to account for parameter uncertainty.

To begin, we need to add parameters that describe our uncertainty in parameters

```{r}
r.sd = 0.2
K.sd = 1.0
NE = 10
```


```{r fig.width=7, fig.height=6}
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}

yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n',ylab="Population Size")
for(i in 1:NE){
  lines(time,n[i,])
}
```

Next if we increase the ensemble size...

```{r,echo=1, fig.width=7, fig.height=6}
NE = 1000
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n',ylab="Population Size")
for(i in 1:NE){
  lines(time,n[i,])
}
hist(rE)
hist(KE)
```

This is hard to visualize, so lets instead look at summary statistics

```{r fig.width=7, fig.height=6}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

```{r fig.width=7, fig.height=6}
n.stats = apply(n,2,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[5, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(n.stats[2, ], rev(n.stats[4, ]),
  n.stats[2, 1])), border = NA, col = 'blue')
lines(time,n.stats[3,],lwd=4,col="purple")
```

Sensitivity analysis at equilibrium

```{r fig.width=9, fig.height=6}
par(mfrow=c(1,2))
plot(rE,n[,NT],ylim=yrng,ylab="Population Size")
rlm <- lm(n[,NT]~rE)
abline(rlm,col=2,lwd=4)
summary(rlm)

plot(KE,n[,NT],ylim=yrng,ylab="Population Size")
Klm <- lm(n[,NT]~KE)
abline(Klm,col=2,lwd=4)
summary(Klm)
```

and at time = 5

```{r fig.width=9, fig.height=6}
par(mfrow=c(1,2))
plot(rE,n[,5],ylim=yrng,ylab="Population Size")
rlm <- lm(n[,5]~rE)
abline(rlm,col=2,lwd=4)
summary(rlm)

plot(KE,n[,5],ylim=yrng,ylab="Population Size")
Klm <- lm(n[,5]~KE)
abline(Klm,col=2,lwd=4)
summary(Klm)
```

and across all times...
```{r fig.width=9, fig.height=6}
rs <- Ks <- rr2 <- Kr2 <- 0
for(i in 1:NT){
  rlm <- lm(n[,i]~rE)
  rs[i] <- coef(rlm)[2]
  rr2[i] <- summary(rlm)$r.squared
  
  Klm <- lm(n[,i]~KE)
  Ks[i] <- coef(Klm)[2]
  Kr2[i] <- summary(Klm)$r.squared
}
par(mfrow=c(1,2))
yrng <- range(c(rs,0))
plot(time,n.stats[3,]/K*yrng[2],lwd=4,col="grey",ylim=yrng,ylab="Sensitivity",type='l')
lines(time,rs,lwd=4)

yrng <- range(c(Ks,0))
plot(time,n.stats[3,]/K*yrng[2],lwd=4,col="grey",ylim=yrng,ylab="Sensitivity",type='l')
lines(time,Ks,lwd=4)

yrng <- range(c(rr2,0))
plot(time,n.stats[3,]/K*yrng[2],lwd=4,col="grey",ylim=yrng,ylab="R-squared",type='l')
lines(time,rr2,lwd=4)

yrng <- range(c(Kr2,0))
plot(time,n.stats[3,]/K*yrng[2],lwd=4,col="grey",ylim=yrng,ylab="R-squared",type='l')
lines(time,Kr2,lwd=4)



```

And sensitivity of MSY
```{r fig.width=7, fig.height=6}
hist(rE*KE/4,xlab="harvest",main="Maximum Sustainable Yield")
```


If K is well known...

```{r,echo=1, fig.width=7, fig.height=6}
K.sd = 0.01
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}
yrng = range(n)
hist(rE)
hist(KE)
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
```

Putting it together
```{r}
r.sd = 0.2
K.sd = 1.0
NE = 5000
nr = nK = n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)
KE = rnorm(NE,K,K.sd)  ## modify later for covariance
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
    nr[i,t] = nr[i,t-1] + rE[i]*nr[i,t-1]*(1-nr[i,t-1]/K)
    nK[i,t] = nK[i,t-1] + r*nK[i,t-1]*(1-nK[i,t-1]/KE[i])
  }
}

layout(matrix(c(1,2,3,4,5,5),3,2,byrow=TRUE))
layout.show(5)

yrng = range(c(range(n),range(nr),range(nK)))
# r
#par(mar=c(1,4,4,2))
par(mar=c(3,4,4,2))
x = seq(min(rE),max(rE),length=500)
plot(x,dnorm(x,r,r.sd),ylab="Density",type='l',
     xlab="Intrinsic Growth Rate",
     cex.lab=1.5,main="r",bty='l',cex.main=3,mgp=c(1.8,0.5,0))
# K
x = seq(min(KE),max(KE),length=500)
plot(x,dnorm(x,K,K.sd),ylab="Density",type='l',
     xlab="Carrying Capacity",
     cex.lab=1.5,main="K",bty='l',cex.main=3,mgp=c(1.8,0.5,0))
# r interval
par(mar=c(3,4,1,2))
nr.stats = apply(nr,2,quantile,c(0.025,0.5,0.975))
plot(time,nr.stats[2,],ylim=yrng,type='n',bty='l',ylab="Population Size",cex.lab=1.5,mgp=c(1.8,0.5,0))
ciEnvelope(time,nr.stats[1,],nr.stats[3,],col="lightGrey")
lines(time,nr.stats[2,],lwd=3)

# K interval
nK.stats = apply(nK,2,quantile,c(0.025,0.5,0.975))
plot(time,nK.stats[2,],ylim=yrng,type='n',bty='l',ylab="Population Size",cex.lab=1.5,mgp=c(1.8,0.5,0))
ciEnvelope(time,nK.stats[1,],nK.stats[3,],col="lightGrey")
lines(time,nK.stats[2,],lwd=3)

# rK interval
par(mar=c(3,4,4,2)+0.1)
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',bty='l',ylab="Population Size",
     cex.main=2,main="r & K",cex.lab=1.5,mgp=c(1.8,0.5,0))
ciEnvelope(time,n.stats[1,],n.stats[3,],col="lightGrey")
lines(time,n.stats[2,],lwd=3)
for(i in 1:10){
  lines(time,n[i,],lty=2)
}
```

Stochastic Population Growth
-------------------------

Lets return to the original model but now create the simplest stochastic version of the model by assuming additive noise

```{r, fig.width=7, fig.height=6}
n0=1
n.sd = 0.25
n = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K) + rnorm(1,0,n.sd)
}
plot(time,n,type='l',ylab="Population Size")
```

```{r fig.width=7, fig.height=6, echo=1}
NE = 1000
n = matrix(n0,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K) + rnorm(1,0,n.sd)
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n',ylab="Population Size")
for(i in 1:10){
  lines(time,n[i,])
}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',bty='l',mgp=c(1.8,0.5,0),
     xlab="Time",ylab="Population Size",cex.lab=1.5)
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
for(i in 1:3){
  lines(time,n[i,],lty=3,lwd=2)
}
n.add = n
```

What if, instead, it is the parameters of the model that vary through time??

```{r fig.width=7, fig.height=6}
r.sd = 0.2
K.sd = 1.0
n = rep(n0,NT)
for(t in 2:NT){  
  rt = rnorm(1,r,r.sd)
  Kt = rnorm(1,K,K.sd)
  n[t] = n[t-1] + rt*n[t-1]*(1-n[t-1]/Kt)
}
plot(time,n,type='l',ylab="Population Size")
```

```{r fig.width=7, fig.height=6,echo=1}
NE = 1000
n1 = n2 = matrix(n0,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    rt = max(c(rnorm(1,r,r.sd),0.001))
    Kt = rnorm(1,K,K.sd)
    n1[i,t] = n1[i,t-1] + r*n1[i,t-1]*(1-n1[i,t-1]/K) + rnorm(1,0,n.sd)
    n2[i,t] = n2[i,t-1] + rt*n2[i,t-1]*(1-n2[i,t-1]/Kt)
  }
}
yrng = range(n2,na.rm=TRUE)
#print(yrng)
plot(time,n2[1,],ylim=yrng,type='n',ylab="Population Size")
for(i in 1:10){
  lines(time,n2[i,])
}
n2.stats = apply(n2,2,quantile,c(0.025,0.5,0.975))
n1.stats = apply(n1,2,quantile,c(0.025,0.5,0.975))
plot(time,n2.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n1.stats[1, ], rev(n1.stats[3, ]),
  n1.stats[1, 1])), border = NA, col = 'lightgreen')
tlb = rgb(173,216,230,100,maxColorValue = 255)
polygon(cbind(c(time, rev(time), 1), c(n2.stats[1, ], rev(n2.stats[3, ]),
  n2.stats[1, 1])), border = NA, col = tlb)
lines(time,n2.stats[2,],lwd=2)
lines(time,n1.stats[2,],lwd=2,col="green")

## two panel version
par(mfrow=c(2,1))
plot(time,n2.stats[2,],ylim=yrng,type='n',
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=2,main="Additive Variability",cex.lab=1.5,mgp=c(1.8,0.5,0))
polygon(cbind(c(time, rev(time), 1), c(n1.stats[1, ], rev(n1.stats[3, ]),
  n1.stats[1, 1])), border = NA, col = 'lightgrey')
lines(time,n1.stats[2,],lwd=3)
for(i in 1:3){
  lines(time,n1[i,],lty=3,lwd=2)
}
plot(time,n2.stats[2,],ylim=yrng,type='n',
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=2,main="Parameter Variability",cex.lab=1.5,mgp=c(1.8,0.5,0))
polygon(cbind(c(time, rev(time), 1), c(n2.stats[1, ], rev(n2.stats[3, ]),
  n2.stats[1, 1])), border = NA, col = tlb)
lines(time,n2.stats[2,],lwd=2)
for(i in 1:3){
  lines(time,n2[i,],lty=3,lwd=2)
}
par(mfrow=c(1,1))
```

what if the parameters are not iid random but autocorrelated?

```{r,echo=1:4, fig.width=7, fig.height=6}
r = 0.3
r.sd = 0.05
K.sd = 1.5
rho = 0.9
NT = 100
time = 1:NT
rt = r; Kt = K
n = rep(n0,NT)
for(t in 2:NT){  
  rt = rnorm(1,r*(1-rho)+rt*rho,r.sd)
  Kt[t] = rnorm(1,K*(1-rho)+Kt[t-1]*rho,K.sd)
  n[t] = n[t-1] + rt*n[t-1]*(1-n[t-1]/Kt[t])
}
plot(time,n,type='l',lwd=3,ylab="Population Size")
lines(time,Kt,col=3,lty=2,lwd=3)
legend("bottomright",legend=c("N","K"),col=c(1,3),lty=1:2,lwd=3)
```

Disturbance
------------------

```{r,echo=1:2, fig.width=7, fig.height=6}
p.dist = 0.05
n0 = 1
r = .3
K = 10
NT = 30
time = 1:NT
NE = 10000
n = matrix(n0,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
    if(rbinom(1,1,p.dist)) n[i,t] = n0
  }
}

plot(time,n[1,],type='l',ylab="Population Size")
plot(time,n[1,],type='l',ylim=range(n),ylab="Population Size")
for(i in 2:min(c(NE,50))){
  lines(time,n[i,])
}

plot(time,n[1,],pch=".",ylim=range(n),ylab="Population Size")
for(i in 2:min(c(NE,100))){
  points(time,jitter(n[i,],amount=0.1),pch=".")
}

hist(n[,NT])
hist(n[,(NT/2):NT],probability=TRUE)

x = n[,(NT/2):NT]
sum(x==n0)/length(x)

yrng = range(n)
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=2,main="Disturbance",cex.lab=1.5,mgp=c(1.8,0.5,0))
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightgrey')
lines(time,n.stats[2,],lwd=3)
for(i in sample(1:NE,5)){
  lines(time,n[i,],lwd=1.5,lty=3)
}

n.dens = matrix(0,NT,K)
N = 1:K
for(i in 1:NT){
  h = hist(n[,i],breaks=0:10,plot=FALSE)
  n.dens[i,] = h$counts
  }
image(time,N,n.dens,col=heat.colors(25))
image(time,N,log(n.dens),col=heat.colors(25))

plot(time,n.dens[,10]/NE,type='l',ylab="Population Size")
for(i in 1:9) lines(time, n.dens[,i]/NE,col=i)
```

Equilibrium solution makes sense. Landscape appears to reach equilibrium almost as fast as undisturbed logistic

Next question, can data assimilation follow this trajectory at the patch or landscape scale?

Initial Conditions
------------------

```{r,echo=1:2, fig.width=7, fig.height=6}
r = 0.3
n0.sd = 0.6
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n',ylab="Population Size")
for(i in 1:10){
  lines(time,n[i,])
}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=1.7,main="Initial Conditions",cex.lab=1.5,mgp=c(1.8,0.5,0))
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightgrey')
lines(time,n.stats[2,],lwd=3)
for(i in sample(1:NE,10)){
  lines(time,n[i,],lwd=1.5,lty=3)
}
```

### sensitivity to initial conditions
```{r, fig.width=7, fig.height=6}
r=2.8
n0=0.1
n = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K)
}
plot(time,n,type='l',ylab="Population Size")
```

```{r,echo=1:2, fig.width=7, fig.height=6}
n0.sd = 0.3
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
plot(time,n[1,],ylim=yrng,type='n',ylab="Population Size")
for(i in 1:10){
  lines(time,n[i,])
}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
hist(n[,NT])
```

```{r,echo=1:2, fig.width=7, fig.height=6}
n0=9
n0.sd = 0.001
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
#plot(time,n[1,],ylim=yrng,type='n')
#for(i in 1:10){
#  lines(time,n[i,])
#}
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=2)
div = n[1,]-n[2,]
```

```{r,echo=1:2, fig.width=7, fig.height=6}
n0=9
n0.sd = 0.03
n0s = rlnorm(NE,log(n0),n0.sd)
hist(n0s)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
yrng = range(n)
#plot(time,n[1,],ylim=yrng,type='n')
#for(i in 1:10){
#  lines(time,n[i,])
#}
n2.stats = n.stats
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(n2.stats[1, ], rev(n2.stats[3, ]),
  n2.stats[1, 1])), border = NA, col = 'lightblue')
lines(time,n.stats[2,],lwd=3,col="green4")
lines(time,n2.stats[2,],lwd=2)
legend("bottomright",legend=c(0.05,0.01),lty=1,col=c("lightgreen","lightblue"),lwd=15,cex=1.5)

layout(matrix(c(1,1,2,3),2,2,byrow=TRUE))
#one ensemble member
r=2.8
n0=9.0001
n = rep(n0,NT)
for(t in 2:NT){
  n[t] = n[t-1] + r*n[t-1]*(1-n[t-1]/K)
}
plot(time,n,type='l',lwd=3,
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=1.7,main="r = 2.8",cex.lab=1.5,mgp=c(1.8,0.5,0))
# divergence

#CI from moderate IC variability
plot(time,n.stats[2,],ylim=yrng,type='n',
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=1.7,main="sd = 0.03",cex.lab=1.5,mgp=c(1.8,0.5,0))
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightgrey')
lines(time,n.stats[2,],lwd=3)

#CI from tiny IC variability
plot(time,n.stats[2,],ylim=yrng,type='n',
     xlab="Time",ylab="Population Size",bty='l',
     cex.main=1.7,main="sd = 0.001",cex.lab=1.5,mgp=c(1.8,0.5,0))
polygon(cbind(c(time, rev(time), 1), c(n2.stats[1, ], rev(n2.stats[3, ]),
  n2.stats[1, 1])), border = NA, col = 'lightgrey')
lines(time,n2.stats[2,],lwd=3)




```

Rapidly diverge to 'climatology', whereby the model prediction does no better than summary statistics

### Foreshadowing: Data Assimilation in a Chaotic System
```{r, fig.width=7, fig.height=6}
r=2.8
K=10
n0=9.01
sd.o = 0.25
n.true = rep(n0,NT)
for(t in 2:NT){
  n.true[t] = n.true[t-1] + r*n.true[t-1]*(1-n.true[t-1]/K)
}
n.obs = rnorm(NT,n.true,sd.o)
plot(time,n.true,type='l',ylab="Population Size")
points(time,n.obs,pch=18)
legend("bottomright",legend=c("True","Obs"),lty=1:0,pch=c(NA,18),lwd=2)
```

```{r,echo=FALSE, fig.width=7, fig.height=6}
n0.sd = 0.01
n0s = rlnorm(NE,log(n0),n0.sd)
ne = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    ne[i,t] = ne[i,t-1] + r*ne[i,t-1]*(1-ne[i,t-1]/K)
  }
}
yrng = range(ne)
n.stats = apply(ne,2,quantile,c(0.025,0.5,0.975))

## initial ensemble
ens = matrix(NA,NT*2+1,NE)
ens[1,] = rlnorm(NE,log(n0),n0.sd)
Xae = Pae = 0
Xfe = mean(ens[1,])
Pfe = var(ens[1,])
for(i in 1:NT){
  
  ##Approximate forecast with Normal
  k = 2*(i-1)+1
  Xfe[i] = mean(ens[k,])
  Pfe[i] = var(ens[k,])
  
  ## analysis step: combine previous forecast with observed data
  H <- matrix(1,ncol=1)
  Kg <- Pfe[i] * t(H) %*% solve(H%*%Pfe[i]%*%t(H) + sd.o)
  Xae[i] <- Xfe[i] + Kg*(n.obs[i] - Xfe[i]) 
  Pae[i] <- (1-Kg %*% H)*Pfe[i]    
  
  ## generate new ensemble
  ens[k+1,] = rnorm(NE,Xae[i],sqrt(Pae[i]))
  
  ## Forecast step: predict to next step from current
  ens[k+2,] = ens[k+1,] + r*ens[k+1,]*(1-ens[k+1,]/K)

#  Xfe[i+1] = m*Xa[i] + beta[2]
#  Pfe[i+1] = q + m*Pa[i]*m
  
}
fx = ens[2*(1:NT)-1,]
fx.stats = apply(fx,1,quantile,c(0.025,0.5,0.975))
plot(time,fx.stats[2,],col="blue",ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(fx.stats[1, ], rev(fx.stats[3, ]),
  fx.stats[1, 1])), border = NA, col = 'lightgreen')
#lines(time,fx.stats[2,],lwd=4,col="green4")
#lines(time,n.stats[2,],lwd=2,col="blue4")
lines(time,n.true,type='l',lwd=2)
points(time,n.obs,pch=18,col=2)
legend("bottomright",legend=c("True","Obs","EnKF","Model"),lty=c(1,0,1,1),col=c(1,2,"lightgreen","lightblue"),lwd=c(1,NA,15,15),cex=1.25,pch=c(NA,18,NA,NA))
```

### Assimilation of disturbance

"True" run

```{r,echo=1:2, fig.width=7, fig.height=6}
p.dist = 0 #0.05
n0 = 1
r = .3
K = 10
n.sd = 0.35
NT = 100
time = 1:NT
NE = 1
n = matrix(n0,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K) + rnorm(1,0,n.sd)
    if(rbinom(1,1,p.dist)) n[i,t] = abs(rnorm(1,n0,0.5))
  }
}

#plot(time,n[1,],type='l',ylab="Population Size")
plot(time,n[1,],type='l',ylim=range(n),ylab="Population Size",lwd=3)
obs.sd = 0.5
n.true=n[1,]
n.obs = rnorm(NT,n.true,obs.sd)
points(time,n.obs,pch=18,col=2,cex=1.5)
na.seq = 1:NT; na.seq=na.seq[-seq(1,NT,by=5)]
n.fia = n.obs;n.fia[na.seq] <- NA
points(time,n.fia,pch=18,col="blue",cex=2)
```

```{r,echo=FALSE, fig.width=7, fig.height=6}
n0.sd = n.sd
NE = 1000
n0s = rlnorm(NE,log(n0),n0.sd)
ne = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    ne[i,t] = abs(ne[i,t-1] + r*ne[i,t-1]*(1-ne[i,t-1]/K) + rnorm(1,0,n.sd))
    if(rbinom(1,1,p.dist)) ne[i,t] = abs(rnorm(1,n0,0.5))
  }
}
yrng = range(ne)
n.stats = apply(ne,2,quantile,c(0.025,0.5,0.975))

## initial ensemble
ens = matrix(NA,NT*2+1,NE)
ens[1,] = rlnorm(NE,log(n0),n0.sd)
Xae = Pae = 0
Xfe = mean(ens[1,])
Pfe = var(ens[1,])
for(i in 1:NT){
  
  ##Approximate forecast with Normal
  k = 2*(i-1)+1
  Xfe[i] = mean(ens[k,])
  Pfe[i] = var(ens[k,])
  
  ## analysis step: combine previous forecast with observed data
  H <- matrix(1,ncol=1)
  Kg <- Pfe[i] * t(H) %*% solve(H%*%Pfe[i]%*%t(H) + obs.sd)
  if(!is.na(n.obs[i])){
    Xae[i] <- Xfe[i] + Kg*(n.obs[i] - Xfe[i]) 
  }else{
    Xae[i] <- Xfe[i]
  }
  Pae[i] <- (1-Kg %*% H)*Pfe[i]    
  
  ## generate new ensemble
  ens[k+1,] = rnorm(NE,Xae[i],sqrt(Pae[i]))
  
  ## Forecast step: predict to next step from current
  ens[k+2,] = ens[k+1,] + r*ens[k+1,]*(1-ens[k+1,]/K) + rnorm(NE,0,n.sd) 
  disturbed = as.logical(rbinom(NE,1,p.dist))
  ens[k+2,disturbed] = abs(rnorm(sum(disturbed),n0,0.5)) ## disturbance

#  Xfe[i+1] = m*Xa[i] + beta[2]
#  Pfe[i+1] = q + m*Pa[i]*m
  
}
fx = ens[2*(1:NT)-1,]
fx.stats = apply(fx,1,quantile,c(0.025,0.25,0.5,0.75,0.975))
plot(time,fx.stats[5,],col="blue",ylim=yrng,type='n',ylab="Population Size",cex.lab=1.5)
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightblue')
polygon(cbind(c(time, rev(time), 1), c(fx.stats[1, ], rev(fx.stats[5, ]),
  fx.stats[1, 1])), border = NA, col = 'lightgreen')
polygon(cbind(c(time, rev(time), 1), c(fx.stats[2, ], rev(fx.stats[4, ]),
  fx.stats[2, 1])), border = NA, col = 'green4')
#lines(time[-NT],fx.stats[2,-1],lwd=4,col="green")
lines(time,n.stats[2,],lwd=2,col="blue4")
lines(time,n.true,type='l',lwd=2)
points(time,n.obs,pch=18,col=2)
legend("topleft",legend=c("True","Obs","95%","IQR","Model"),lty=c(1,0,1,1,1),col=c(1,2,"lightgreen","green4","lightblue"),lwd=c(1,NA,15,15,15),cex=1.2,pch=c(NA,18,NA,NA,NA))
```

```{r,echo=1:2, fig.width=7, fig.height=6}
p.dist = 0.05
n0 = 1
r = .3
K = 10
NT = 30
time = 1:NT
NE = 1000
n = matrix(n0,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
    if(rbinom(1,1,p.dist)) n[i,t] = n0
  }
}

plot(time,n[1,],type='l',ylab="Population Size")
plot(time,n[1,],type='l',ylim=range(n),ylab="Population Size")
for(i in 2:min(c(NE,50))){
  lines(time,n[i,])
}

plot(time,n[1,],pch=".",ylim=range(n),ylab="Population Size")
for(i in 2:min(c(NE,100))){
  points(time,jitter(n[i,],amount=0.1),pch=".")
}

hist(n[,NT])
hist(n[,(NT/2):NT],probability=TRUE)

x = n[,(NT/2):NT]
sum(x==n0)/length(x)

n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
plot(time,n.stats[2,],ylim=yrng,type='n',ylab="Population Size")
polygon(cbind(c(time, rev(time), 1), c(n.stats[1, ], rev(n.stats[3, ]),
  n.stats[1, 1])), border = NA, col = 'lightgreen')

n.dens = matrix(0,NT,K)
N = 1:K
for(i in 1:NT){
  h = hist(n[,i],breaks=0:10,plot=FALSE)
  n.dens[i,] = h$counts
  }
image(time,N,n.dens,col=heat.colors(25))
image(time,N,log(n.dens),col=heat.colors(25))

plot(time,n.dens[,10]/NE,type='l',ylab="Population Size")
for(i in 1:9) lines(time, n.dens[,i]/NE,col=i)
```
